---
title: "Biomarkers of ASD"
subtitle: "If you want a subtitle put it here"
author: "Andrew Guerra, "
date: last-modified
published-title: "Updated"
editor: visual
format: html
code-copy: true
execute:
  message: false
  warning: false
  echo: false
  cache: true
---

Use this as a template. Keep the headers and remove all other text. In all, your report can be quite short. When it is complete, render and then push changes to your team repository.

```{r, message=FALSE}
# load any other packages and read data here
library(tidyverse)
library(here)
library(readr)
library(knitr)
raw.biomarkers <- read_csv(here("data", "biomarker-raw.csv"))

```

## Abstract

Write a brief one-paragraph abstract that describes the contents of your write-up.

## Dataset

Write a brief data description, including: how data were obtained; sample characteristics; variables measured; and data preprocessing. This can be largely based on the source paper and should not exceed 1-2 paragraphs.

## Summary of published analysis

Summarize the methodology of the paper in 1-3 paragraphs. You need not explain the methods in depth as we did in class; just indicate what methods were used and how they were combined. If possible, include a diagram that depicts the methodological design. (Quarto has support for [GraphViz and Mermaid flowcharts](https://quarto.org/docs/authoring/diagrams.html).) Provide key results: the proteins selected for the classifier and the estimated accuracy.

## Findings

Summarize your findings here. I've included some subheaders in a way that seems natural to me; you can structure this section however you like.

### Impact of preprocessing and outliers

**What do you imagine is the reason for log-transforming the protein levels in biomarker-raw.csv?**

Before protein levels are measured, a log-transformation is applied. The protein levels in biomarker-raw.csv are logarithmically transformed to better capture small variations in biomarker level expression. When working with microscopic data such as protein levels, the range of values is often dramatic. As a result, data can often be accentuated more or less and make overall trends harder to detect. To control for this, we can apply a log-transformation to standardize our data range.

To verify this, we can sample from 5 proteins in the dataset and see their respective distributions.

```{r, out.width='80%', fig.align='center'}
set.seed(10302025)
# Change Headers to Protein Acronyms
raw.biomarkers1 <- raw.biomarkers
colnames(raw.biomarkers1) <- raw.biomarkers1[1,]
raw.data <- raw.biomarkers1[-1,]
sample.proteins <- sample(colnames(raw.data), 5)
# RS of n=4 data
cat('Our proteins sampled are: ', sample.proteins)

# Checking distributions
long.data <- raw.data %>%
  select(all_of(sample.proteins)) %>%
  pivot_longer(cols = everything(), names_to = "Protein", values_to = "Value") %>%
  mutate(Value = as.numeric(Value))

# Histogram
long.data %>% 
  drop_na() %>% 
ggplot(aes(x = Value)) +
  geom_histogram(color = "white", fill = "black", bins = 30) +
  facet_wrap(~ Protein, scales = "free_x") +
  labs(
    title = "Distributions of Selected Protein Biomarkers",
    x = "Protein Level",
    y = "Frequency"
  ) +
  theme_minimal()

```

After plotting the histograms of our sampled proteins, we see that they most (except Apo E4) exhibit notable skewness and are not normally distribution. Additionally, the range for these values is extremely wide. Next, we apply a log-transformation to help standardize our distribtutions.

```{r, fig.align='center', out.width='80%'}
set.seed(10302025)

long.data %>% 
  drop_na() %>% 
ggplot(aes(x = log(Value))) +
  geom_histogram(color = "white", fill = "black", bins = 30) +
  facet_wrap(~ Protein, scales = "free_x") +
  labs(
    title = "Distributions of Selected Protein Biomarkers",
    x = "Log-Transformed Protein Level",
    y = "Frequency"
  ) +
  theme_minimal()
```

After transforming our values, we see that the range is now much considerably smaller. Additionally, most of our proteins sampled follow a close-to-normal distribution. Interestingly, Coagulation Factor IX appears to have a median of about 8.8. This trend was not as apparent with applying a transformation. Thus, we have reason to suspect that this protein should be further investigated.

**Temporarily remove the outlier trimming from preprocessing and do some exploratory analysis of the outlying values. Are there specific subjects (not values) that seem to be outliers? If so, are outliers more frequent in one group of the other?**

Through our previous analyses, we see that the distribution of proteins are transformed to easily detect which have a non-normal relationship that can be investigated further. Shifting our focus, we aim to see whether subjects are outliers. That is, we aim to see whether there are subjects, portrayed as rows, that consist of outliers among multiple protein levels.

```{r fig.align='center', warning=FALSE, out.width='80%'}
proteins <- setdiff(names(raw.biomarkers), 
                    c("Group", "Target Full Name"))

z.scores <- raw.biomarkers %>% 
  mutate(across(all_of(proteins), ~scale(as.numeric(.x))))


outlier_summary <- z.scores %>% 
  mutate(across(all_of(proteins), ~abs(.x) > 3)) %>% 
  mutate(n_outliers = rowSums(across(all_of(proteins)), na.rm = TRUE)) %>% 
  filter(!is.na(Group) & Group != '') %>%
  group_by(Group) %>% 
  summarise(
    mean_outliers = mean(n_outliers, na.rm = TRUE),
    median_outliers = median(n_outliers, na.rm = TRUE),
    sd_outliers = sd(n_outliers, na.rm = TRUE),
    max_outliers = max(n_outliers, na.rm = TRUE),
    .groups = "drop"
  )

outlier_summary %>% kable(caption = 'Outlier Distribution Table', digits=2)
```

We created a table to see outlier trends by subject group. To do so, we standardized all 1,317 protein levels and counted values with an absolute value greater than 3 as an outlier. Based on our table, we see that ASD subjects had an average of about 17.04 outliers in protein levels while TD subjects had about 19.65. Additionally, the median outlier counters were nearly identical (9.5 for ASD and 10 for TD), indicating that the two groups have similar overall variation in protein measurements. However, the standard deviations (20.6 for ASD and 30.4 for TD) and maximum outlier counts (119 and 154, respectively) suggest that a few individual subjects in both groups exhibited unusually high numbers of outlying protein values. Overall, this does not suggest there being systematic differences between groups, rather it suggests there are few subjects in the data that have greater individual differences.

### Methodological variations

Task 3

### Improved classifier

Task 4
